Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-18 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Merge Lisa and Moritz
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2878 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 863de85c66db6ac7a6aae305563de0c493feaf43

The corresponding log-likelihood function is defined by
\begin{equation}
\begin{aligned}
\ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
&= -\log(2 \pi I_0(\kappa)) +  \kappa \cos(y - \mu).
\end{aligned}
\end{equation}

@@

\subsection{Data}
As response variables, $10$\,min mean wind direction observations are employed at
the airports Innsbruck and Vienna on an hourly temporal resolution. As
predictor variables, we use observations of various meteorological quantities
such as wind direction, temperature, air pressure and humidity directly at the
respective airports, as well as within the vicinity of the airports. For
Innsbruck, meteorological variables are observed at XXX different locations
along the intersecting valleys; whereas for Vienna, located in the flat,
meteorological variables are observed at XXX different locations within an area
of about $10$\,km$^2$. In addition, we employ temporal differences, averages, minima, maxima over the last 3\,h,

of ias well as tempat each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
especially of the employed reference models might suffer from the influence of these ``noise'' 
observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
considered.

\fixme{information on NAs, update data information, include overview}

@@

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular regression trees and forest and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches in the field of meteorological forecasting while the
third is a state-of-the-art model to forecast circular response variables. The
next paragraphs introduce the reference methods and provide the corresponding
model specifications as employed in this study for all considered model
including the circular tree and forest.

\paragraph{Circular regression tree}
For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
learning data are considered. The tree is obtained by employing the CTree algorithm with a minimal 
number of 2000 observations in each terminal node serving as a stopping criterion 
(argument \code{minbucket} of our implementation).

\paragraph{Circular regression forest}
Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
As for the distributional tree all available covariates are used to learn a distributional 
forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
containing a fraction of 30 percent of the original learning data. As it can be expected that 
the lagged response variable (as included as a covariate) has a strong influence on the 
response variable it is desirable to employ it in each tree of the forest. Therefore, all 
covariates are considered for each tree
(by setting the argument \code{mtry} of our implementation to the total number of covariates)
in order to guarantee for the lagged response to be available.
Since a high number of possible split points leads to a high computational effort that is 
particularly noteworthy for forest models the covariates are binned allowing for a maximum of
$50$ classes. In that way the number of split points that have to be considered is reduced
in a reasonably.
% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
maximum value of 1 in order to build trees as large as possible as commonly done for forest models.

\paragraph{Persistency model}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.

\fixme{add references}

To establish a distributional model of persistence in this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the six previous time points are considered. In order to allow for a stronger influence of 
observations closer to the prediction time point exponential smoothing is employed, i.e.,
exponentially decaying weights are included with a smoothing factor of $0.5$ such that
in each prediction step the current prediction and the previous observation both have an 
equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
For this selected set of weighted observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the 
circular CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.
\fixme{not done like this, always using the whole data set. wrong?}

\paragraph{Climatological model}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected time point and day of the year weather conditions will always be the same over all 
years considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.

\fixme{add references}

Similar to the applied persistence model a set of observations is selected to fit a distributional
climatology model. For a specified day and time point observations from the exact same time point of 
the same day as well as of the 30 surrounding days from all other years contained in the learning data 
are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.

\paragraph{Generalized linear model}
Another commonly applied approach is to linearly regress the circular variable of interest on
its lagged values. For the here discussed application this corresponds to establishing a generalized
linear model with wind direction as the outcome variable and wind direction of the previous time point 
as covariate. 
Following \cite{Mulder+Klugkist:2017} the employed model plugs in a link function in order
to guarantee values on the interval $[-\pi,\pi]$ and is of the form
\begin{align}
dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
\end{align}
with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
The intercept and the regression coefficient are estimated in a Bayesian framework where an 
MCMC algorithm is used to obtain posterior samples. Moreover, a constant concentration
parameter is fitted to the full learning sample.

%\fixme{add references and further information on the estimation process within circGLM?}

%\fixme{add information on applied implementation here (!) or in separate chapter?}

>>>>

The corresponding log-likelihood function is defined by
\begin{equation}
\begin{aligned}
\ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
&= -\log(2 \pi I_0(\kappa)) +  \kappa \cos(y - \mu).\label{equ:vM:loglik}
\end{aligned}
\end{equation}

@@

\subsection{Data}\label{sec:wind:data}
For the case study, as circular response variables we employ $10$\,min mean
wind direction measurements at the airports Innsbruck and Vienna on an hourly
temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all
lagged by one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
the vicinity of the respective airports. For Innsbruck, measurements of
21~different locations at the airport and along the intersecting valleys are
used, whereas, for Vienna, measurements of 22~locations at the airport and
within an area of about $10$\,km$^2$ are used. A topographical overview of the
airports and their surrounding areas with the station sites employed in this
study is provided in Figure~XXX. In addition, we use derived quantities such
as 3~hourly means, minima and maxima, as well as 1~and 3~hourly temporal
changes and spatial differences towards the airport of the respective
quantities. A summary of all predictor variables can be found in
Table~\ref{tab:covariates}.

The data used in this study consists of 5~total years from January 2015 to
December 2018. After first eliminating predictor variables with more than 5\% missing
values and then cases with missing observations, the
data set consists of 41979 time points and 260 covariates for Innsbruck, and of
38985 times points and 494 covariates for Vienna, respectively.

@@

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular regression trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches in the meteorological field while the third is a
state-of-the-art model to forecast circular response variables. The next
paragraphs introduce the reference methods and provide the corresponding model
specifications as employed in this study for all considered models including the
circular tree and forest.

\paragraph{Climatological model} 
Accurate knowledge of weather quantities' climatologies can be important for a
wide range of applications and are often also used as a baseline for the
validation of newly developed forecasting systems
\citep{simon.etal:2017,stauffer.etal:2017}. Usually, the performance of
climatological models strongly depends on the temporal dependency of the
quantity itself, either on a daily or yearly basis. Wind direction at a single
location is often quite unstable and shows only a daily or annual
seasonality in case of predominant local wind systems like the ones at
Innsbruck introduced for the analysis of Figure~\ref{fig:tree_ibk}. Hence,
the climatological model in this study provides a very simple implementation of
a baseline model and might experience serious performance losses in case of
random temporal variabilities or abrupt changes in the employed response
variable wind direction.

For simplicity, rather to fit one single climatological model, we perform the
model fitting for each time point separately with varying training data. For
each time point, the training data set consists of the same daytime as the time
of interest within 31~days centered around the same calender day over all years
except the year of interest. To gain full probabilistic climatological
forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
the distribution parameters of the von Mises distribution via maximum
likelihood estimation. The construction of a climatological reference model
was similarly pursued by \citet{vogel.etal:2018} and is discussed in a
comprehensive summary on different time-adaptive training scheme in
\citet{lang.etal:2019c}.

\paragraph{Persistency model} 
Persistence describes the previous value of a single weather quantity in a time
series; hence, it is another objective measure in the verification of
weather forecasts and also often applied as a reference model
\citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
atmospheric conditions, persistence can be a very good estimate, but as soon as
any major change takes place in the response variable during the forecast
period, forecasts based on persistence may also fail remarkably. Therefore, it
heavily depends on the variability of the variable of interest and the length
of the forecasting step whether persistence can provide overall reliable
predictions.
%If a certain weather quantity remains constant over a certain time period,
%persistence can be a very good estimate.

To gain a full probabilistic persistency model, we use as for the
climatological model maximimum likelihood estimation and fit the distribution
parameters of the von Mises distribution conditional on lagged response values
according to the description in Section~\ref{sec:prob_circ}. We fit one model
for every time point in the validation data set employing the previous six
lagged response values as training data. In order to allow for a stronger
influence of observations closer to the time of interest an exponential
smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
prediction the current observation and all previous five observations together
have an equal influence rate of 50 percent. Observations with longer time lags
have exponential weights below $0.01$ and are therefore omitted
from the training data.

\paragraph{Generalized linear model}
Traditional approaches to forecast circular response variables are often based on
circular GLMs \citep{Fisher:1993}. As discussed in
Section~\ref{sec:introduction}, circular regression models often experience
the problem that the likelihood function can be strongly irregular which makes
optimization rather difficult. Hence, they often do not converge if no
appropriate initial values are provided \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In this study, to be able to employ a GLM out of the box
as a reference, we use the Bayesian implementation of
\cite{Mulder+Klugkist:2017} which does not dependent so much on initial values
due to an MCMC sampling algorithm using weakly- or non-informative priors.

Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
appropriate link function~$g$ in order to guarantee response values on the
interval $[-\pi,\pi]$. As the implementation cannot handle circular covariates,
we use the components of the lagged 2-dimensional wind vector $vec$ and the lagged
wind speed $spd$ as predictor variables. The model formula for the location parameter
$mu$ of the von Mises distribution can be written as:
%%
\begin{align}
\mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd
\end{align}
%%
with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In addition, a
constant concentration parameter $\kappa$ is fitted to the full learning sample.

\paragraph{Circular regression tree}
For the circular regression tree introduced in Section~\ref{sec:circtree}, all
covariates provided in the learning data, as described in
Section~\ref{sec:wind:data}, can be considered due to an intrinsic automatic
variable selection performed in the tree estimation. The tree building is
performed by our newly developed R-package \pkg{circtree} employing the CTree
algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a minimal number of 2000
observations in each terminal node serving as a stopping criterion (argument
\code{minbucket}).

\paragraph{Circular regression forest} Following the description in
Section~\ref{sec:circforest}, a circular regression forest is constructed based
on 100 individual trees employing our R-package \pkg{circtree}. Each of these
trees is again built by the CTree algorithm on a subsample containing a
fraction of 30 percent of the original learning data. As it can be expected
that the lagged response variable is one of the most informative predictor
variables for wind direction forecasting, all covariates are considered for
building each tree. This `bagging' ensures that the lagged response
variable is included in each tree; in our implementation this can be done by
setting the argument \code{mtry} to the total number of covariates. Since a
high number of possible split points leads to high computational costs, the
covariates are automatically binned in a maximum of $50$ classes (argument
\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to building a single tree,
forests usually consists of large trees as they are not prone to overfitting
the data after some kind of averaging over the individual trees. Therefore, we
use the following control arguments to build rather long trees: The minimal number 
of observations to perform a split is set to 20
(argument \texttt{minsplit}), the minimal number of observations in each
terminal node is set to 7 (argument \texttt{minbucket}), and the significance
level for variable selection is kept at its maximum value of $1$
(argument \texttt{alpha}).
